sdmTMB: Fast, flexible, and user-friendly spatial and spatiotemporal generalized linear mixed-effects models

### 2021-11-27 doc/paper.Rmd
Major re-arrange

## Model selection

Following glmmTMB, we allow AIC to be calculated for each sdmTMB model [`aic()`, @akaike1974]. 
Because of well-documented biases of AIC with mixed-effects models [@liang2008] we provide functionality to perform k-fold cross-validation.
The function `sdmTMB_cv()` allows for any number of folds, with assignments being user-specified or chosen randomly. 
The predictive log-density is calculated for each fold and summarized across folds for a total measure of the predictive accuracy of a model.

## Model evaluation

Validation of mixed-effects models is non-trivial (that REF). We provide three approaches to assist with model validation:

1. The `residuals()` method returns randomized quantile residuals (e.g., as in DHARMa REF).
These residuals are calculated as $\hat{y} - y$ where $\hat{y}$ is the sum of the fixed effect estimates and the random effect values that maximumize the likelihood conditional on the estimated fixed effects. 
These residuals have known statistical issues (REF) but are quick to calculate. One-step-ahead residuals (REF) or residuals from a single MCMC draw (REF) are also possible, but both approaches are much slower.

2. The `sdmTMB_simulate()` function can simulate new data, or simulate from a fitted model, to which models can be fit to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.

3. An sdmTMB model can be passed to the tmbstan package [@monnahan2018] to sample from the joint posterior with Stan (REF) to evaluate the accuracy of the Laplace approximation (see `?extract_mcmc`). 
We note this will be considerably slower to fit than with maximum likelihood and in some cases may require tight priors or fixed parameters to ensure adequate sampling (REF).

The results from an sdmTMB model can be used to generate various quantities of interest.
For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity (`get_cog()`) and total population index (`get_index()`).
Estimated parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range.

@@

We could generate a standardized index from that fit on a 2x2 km grid by supplying the `area` per grid cell and telling `predict.sdmTMB()` to return a necessary TMB object (Fig. \@ref(fig:pcod-fig)E). `get_index()` sums the biomass across grid cells to derive an area-weighted index of biomass independent of sampling locations.

```{r, eval=FALSE, echo=TRUE}
p <- predict(fit_spatiotemporal, newdata = qcs_grid, area = 4, return_tmb_object = TRUE)
ind <- get_index(p)
```

@@

(ref:owl-nao-fig) Spatially varying coefficient for effect of mean annual NAO on counts of Snowy Owls observed on annual Christmas Bird Counts 1979--2020. 
Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8). 
The effect is multiplicative on owl count per NOA unit.

@@

@@

(ref:timing-fig) Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions.
Lines represent means and ribbons ranges across 20 random iterations.
Note the log x and y axes.
sdmTMB used the default settings and `normalize = TRUE`, which calculates the random effect normalizing constant in R.
inlabru used the empirical Bayes integration strategy and Gaussian approximation with `bru_max_iter = 1`, and the `like()` formulation.
mgcv used `bam()`, `method = fREML`, and discretized covariates [@miller2019].
inlabru/INLA could be faster with the PARDISO sparse linear solver and TMB could be faster with an optimized BLAS library.
All platforms were restricted to one core and could be faster with parallel computation.
<!-- https://www.sciencedirect.com/science/article/abs/pii/S0167739X00000765 -->

@@

There are many R packages available for fitting spatially explicit models to large datasets and we cannot review them all here [see @heaton2017]; however, the three most closely related are VAST, inlabru, and mgcv (Table 1). 
VAST is also built on TMB and, for univariate data, should be able to fit equivalent models with some minor differences in functionality.
The difference lies mainly in the user interfaces and considerable additional functionality for multivariate response data (REF) and delta/hurdle models (REF) in VAST.
Interface-wise, VAST is a less modular interface (e.g., by default automatically constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries index standardization perspective (e.g., splitting 'catchability' and 'habitat' covariates), and lacks some interface elements that we think make sdmTMB more broadly usable by applied ecologists. 
For example, `sdmTMB()` has a `family` argument as in `glm()`, a `predict()` method, and human-readable options for spatial/spatiotemporal fields (e.g., `spatial = 'on'`, `spatiotemporal = 'AR1'`).
sdmTMB achieves three of the four design principles outlined in @thorson2019 for index standardization: it allows area weighting, can distinguish catchability and habitat covariates (by deciding what is fixed when predicting), and conditions on missing covariates (random fields).
Unlike VAST, however, sdmTMB does not currently extend to multivariate applications.

>>>>

@@

@@

(ref:owl-nao-fig) Spatially varying coefficient for effect of mean annual NAO (North Atlantic Oscillation) on counts of Snowy Owls observed on annual Christmas Bird Counts 1979--2020 in Canada and the US. 
Points represent all count locations and circle area is scaled to the mean number of owls observed per year (range: 0 to 8). 
The effect is multiplicative on owl count per NOA unit.

@@

## Model selection and validation

Following glmmTMB, we allow AIC to be calculated for each sdmTMB model [`AIC()`, @akaike1974]. 
Because of well-documented biases of AIC with mixed-effects models [@liang2008] we provide functionality to perform k-fold cross-validation.
The function `sdmTMB_cv()` allows for any number of folds, with assignments being user-specified or chosen randomly. 
The predictive log-density is calculated for each fold and summarized across folds for a total measure of the predictive accuracy of a model.

Validation of state-space models is non-trivial, particularly those using the Laplace approximation [@thygesen2017a]. We provide four approaches to assist with model validation:

1. The cross validation function `sdmTMB_cv()` can be used to evaluate goodness of fit according to user-calculated criteria (e.g., mean squared error, area under the curve).

2. The `residuals()` method returns randomized quantile [@dunn1996] or probability integral transform (PIT) [@smith1985] residuals.
These residuals, however, are calculated as $\hat{y} - y$ where $\hat{y}$ is the sum of the fixed effect estimates and the random effect values that maximize the likelihood conditional on the estimated fixed effects. 
These residuals have known statistical issues [@thygesen2017a] but are quick to calculate.
One-step-ahead residuals [@thygesen2017a] or residuals from a single MCMC draw [e.g., @rufener2021] are also possible, but are much slower.

3. The `simulate.sdmTMB()` method or `sdmTMB_simulate()` function can simulate from fitted models or new data to which models can be fit to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.
Simulations can also be passed to the R package DHARMa [@DHARMa] to calculate simulation-based residual diagnostics (APPENDIX OR VIGNETTE).

4. An sdmTMB model can be passed to the tmbstan package [@monnahan2018] to sample from the joint posterior with Stan [@carpenter2017] to evaluate the accuracy of the Laplace approximation or evaluate residuals (see `?extract_mcmc`). 
<!-- We note this will be considerably slower to fit than with maximum likelihood and in some cases may require tight priors or fixed parameters to ensure adequate sampling [as in @monnahan2021]. -->

The results from an sdmTMB model can be used to generate various quantities of interest.
For example, sdmTMB provides functionality for generating population-level summaries for each time slice including the center of gravity (`get_cog()`) and total population index (`get_index()`; Fig. \@ref(fig:pcod-fig)E).
Estimated parameters from the Matérn covariance function can be used to describe the marginal spatial variance and Matérn range.

<!-- We could generate a standardized index from that fit on a 2x2 km grid by supplying the `area` per grid cell and telling `predict.sdmTMB()` to return a necessary TMB object (Fig. \@ref(fig:pcod-fig)E). `get_index()` sums the biomass across grid cells to derive an area-weighted index of biomass independent of sampling locations. -->

```{r, eval=FALSE, echo=FALSE}
p <- predict(fit_spatiotemporal, newdata = qcs_grid, area = 4, return_tmb_object = TRUE)
ind <- get_index(p)
```

@@

(ref:timing-fig) Comparison of time to fit an SPDE spatial random field model with 1000 observations, an intercept and one predictor, Gaussian error, and a sequence of SPDE resolutions.
Lines represent means and ribbons ranges across 20 random iterations.
Note the log x and y axes.
sdmTMB used the default settings and `normalize = TRUE`, which calculates the random effect normalizing constant in R.
inlabru used the empirical Bayes integration strategy and Gaussian approximation with `bru_max_iter = 1`, and the `like()` formulation.
mgcv used `bam()`, `method = fREML`, and discretized covariates [@miller2019].
<!-- inlabru/INLA could be faster with the PARDISO sparse linear solver and TMB could be faster with an optimized BLAS library. -->
<!-- All platforms were restricted to one core and could be faster with parallel computation. -->
<!-- https://www.sciencedirect.com/science/article/abs/pii/S0167739X00000765 -->

@@

There are many R packages available for fitting spatially explicit models to large datasets and we cannot review them all here [see @heaton2017]; however, the three most closely related are VAST, inlabru, and mgcv (Table 1). 
VAST is also built on TMB and, for univariate data, should be able to fit equivalent models with some minor differences in functionality.
The difference lies mainly in the user interfaces and considerable additional functionality for multivariate response data (REF) and delta/hurdle models (REF) in VAST.
Interface-wise, VAST is a less modular interface (e.g., by default constructing meshes before fitting, making predictions at the same time as fitting), is written from a fisheries perspective, and lacks some interface elements that we think make sdmTMB more broadly usable by applied ecologists. 
For example, `sdmTMB()` has a `family` argument as in `glm()`, a `predict()` method, and human-readable options for spatial/spatiotemporal fields (e.g., `spatial = 'on'`, `spatiotemporal = 'AR1'`).
sdmTMB achieves three of the four design principles outlined in @thorson2019 for index standardization: it allows area weighting, can distinguish catchability and habitat covariates (by deciding what is fixed when predicting), and conditions on missing covariates (random fields).
Unlike VAST, however, sdmTMB does not currently extend to multivariate applications.
